Automatically Proving Equivalence by Type-Safe Reflection

### 2017-05-19 paper-CICM-17/content/1-introduction.tex
Some additional changes

Proofs assistants like Coq~\cite{BertotC04} 
and programming languages like 
Agda~\cite{DBLP:conf/tldi/Norell09} and
Idris~\cite{brady2013idris} are based on Intensional Type Theories that contain
two notions of equality: propositional equality, that can be
manipulated in the language, and judgemental (or definition) equality.
Propositional equality corresponds to the mathematical notion: this is
a proposition that can be assumed, negated, proved or disproved. Since in type
theory, propositions are types~\cite{How80}, the proposition that two
elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parameterised by the type $a$) indexed over two elements of $a$, giving $\code{Id} (a:\code{Type}):a \rightarrow a \rightarrow \code{Type}$. For
convenience, we write $(\code{Id}_a\ x\ y)$ as $(x\ =_a\ y)$.

@@

The only way for $(\code{Id}_a\ x\ y)$ to be inhabited is therefore that $x$
equals $y$. In this case, the constructor \code{Refl} helps
to create a proof of this equality: \code{(Refl\ x)} is precisely the proof
which says that $x=_ax$.

@@

We will write a function to add two binary numbers.
To do so, we begin with an auxiliary function, which adds three bits
(the third is a carry bit), and produces the two bits of
the result, where the first is the more significant bit:

@@

Unfortunately, this definition is rejected because the types of each side do
not match.  The result of the first line \code{adc zero zero carry} is expected
to have the type:

\codeL{Binary\ 1\ ((c\ + 0)\ + 0)} but we provide a
term of type:

\code{Binary\ 1\ (c\ + (0\ + (0\ + 0)))}.

The problem is even worse for the second case where the expected index is:

@@

Such proofs consist of a potentially long sequence of
rewriting steps, each using one of the properties: neutral element,
commutativity, associativity. Without some automation, this sequence of
rewritings must be done by the programmer.  Not only is this time consuming,
but a small change in the definition may lead to a different proof obligation,
thus invalidating the proof. 
A minor change in the datatype, or 
the definition of \code{addBit} or \code{adc} will require us
to do a new proof, and thus, without support from the machine, these
proofs
could become the everyday routine in any dependently-typed language.

@@

Provers for some algebraic structures have already been implemented for various
proof assistants, including Coq~\cite{Coq2005} and
Agda\footnote{\url{http://wiki.portal.chalmers.se/agda\%5C?n=Libraries.UsingTheRingSolver}}. 
In this paper, we describe an implementation\footnote{The
implementation of our hierarchy of tactics can be found online at
\url{https://github.com/FranckS/RingIdris/Provers}} of an automatic prover for
equalities in a \emph{hierarchy} of algebraic structures, including monoids,
groups and rings (all potentially commutative), for the Idris language.,
making the following contributions:

>>>>

Proofs assistants like Coq~\cite{BertotC04} 
and programming languages like 
Agda~\cite{DBLP:conf/tldi/Norell09} and
Idris~\cite{brady2013idris} are based on Intensional Type Theories that contain
two notions of equality: propositional equality, that can be
manipulated in the language, and judgemental (or definitional) equality.
Propositional equality corresponds to the mathematical notion: this is
a proposition that can be assumed, negated, proved or disproved. Since in type
theory, propositions are types~\cite{How80}, the proposition that two
elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parameterised by the type $a$) indexed over two elements of $a$, giving $\code{Id} (a:\code{Type}):a \rightarrow a \rightarrow \code{Type}$. For
convenience, we write $(\code{Id}_a\ x\ y)$ as $(x\ =_a\ y)$.

@@

The only way for $(\code{Id}_a\ x\ y)$ to be inhabited is therefore that $x$ and $y$ are equal by definition. In this case, the constructor \code{Refl} helps to create a proof of this equality: \code{(Refl\ x)} is precisely the proof
which says that $x=_ax$.

@@

We will write a function to add two binary numbers.
To do so, we begin with an auxiliary function, which adds three bits
(the third is a carry bit), and produces the two bits of
the result, where the first is the most significant bit:

@@

Unfortunately, this definition is rejected because the types do
not match for both patterns.  
%The result of the first line \code{adc zero zero carry} is expected
%to have the type:
%
%\codeL{Binary\ 1\ ((c\ + 0)\ + 0)} but we provide a
%term of type:
%
%\code{Binary\ 1\ (c\ + (0\ + (0\ + 0)))}.
For the second case, the expected index is:

@@

Such proofs consist of a potentially long sequence of
rewriting steps, each using one of the properties: neutral element,
commutativity, associativity. Without some automation, this sequence of
rewritings must be done by the programmer.  Not only is this time consuming,
but a small change in the definition may lead to a different proof obligation,
thus invalidating the proof. 
A minor change in the datatype, or 
the definition of \code{addBit} or \code{adc} will require us
to do a new proof, and thus, without support from the machine, these
proofs
could become the everyday routine in any dependently-typed language. It is worth mentioning that even without using dependent types, these proof obligations for equalities happen very frequently during the formal certification of most applications.

@@

Provers for some algebraic structures have already been implemented for various
proof assistants, including Coq~\cite{Coq2005} and
Agda\footnote{\url{http://wiki.portal.chalmers.se/agda\%5C?n=Libraries.UsingTheRingSolver}}. 
In this paper, we describe an implementation\footnote{The
implementation of our hierarchy of tactics can be found online at
\url{https://github.com/FranckS/RingIdris/Provers}} of an automatic prover for
equalities in a \emph{hierarchy} of algebraic structures, including monoids,
groups and rings (all potentially commutative), for the Idris language,
making the following contributions:

### 2017-05-19 paper-CICM-17/content/3-hierarchyOfTactics.tex
Some additional changes

With some additional effort, we can produce a collection of tactics for proving
\emph{equivalences}, rather than only equalities. 
The machinery is very similar and
we gain another
degree of genericity, with the freedom of choosing the equivalence relation
(which can be the usual equality). 
The user can define their own
notion of equivalence, as long as they provide the proofs of the
properties of the relevant algebraic structure. Let's call \code{c} the
\emph{carrier} type, i.e., the type on which we want to prove equivalences. 
The
equivalence relation on \code{c} has the following profile $(\simeq) : \code{c}
\rightarrow \code{c} \rightarrow \code{Type}$\footnote{This \code{Type} would
be a \code{Prop} in systems, like Coq, that make a distinction between the
world of computations and the world of logical statements}, and must be
be reflexive, symmetric and transitive.

@@

\begin{lstlisting}
interface Set c where
    ($\simeq$) : c $\rightarrow$ c $\rightarrow$ Type
    refl : (x : c) $\rightarrow$ x $\simeq$ x
    sym : {x, y : c} $\rightarrow$ (x $\simeq$ y) $\rightarrow$ (y $\simeq$ x)
    trans : {x, y, z : c} $\rightarrow$ (x $\simeq$ y) $\rightarrow$ (y $\simeq$ z) $\rightarrow$ (x $\simeq$ z)    
    set_eq : (x : c) $\rightarrow$ (y:c) $\rightarrow$ Maybe (x $\simeq$ y)
\end{lstlisting}

@@

Working with equivalences instead of equalities brings one complication: Proofs
of correctness that we produce by hand cannot use Idris' ``rewrite'' mechanism,
which enables rewriting of a subterm by another one, provided that the two
subterms are propositionally equal. 
%
% EB - I think we can skip this detail
%If we need to prove $P\ x$,
%and have a proof of equivalence $pr:x \simeq x'$, we can't use ``rewrite pr" in
%order to transform the goal into $P\ x'$, as we would normally do if the
%available proof would be an equality $x=x'$. 
%
This is a classical problem of
working within a setoid, which can be mitigated by programming
language support for rewriting terms in setoids. However, Idris
is not equipped with any such support. For this reason, we define the 
following lemma, using the methods of the \code{Set} interface:

@@

We have this requirement because we support
any equivalence relation.
As with \code{Set}, there is no
tactic for \code{Magma}, because there is no property; all
equivalences are again syntactic equivalences, and can thus be proven by
\code{refl}.

@@

\noindent
Examples of magma are \code{Nat} equipped with addition, and \code{List} with
concatenation.  Next, a monoid is a semi-group with a ``neutral``, or
distinguished, element called \code{Zero}.

@@

When proving an equivalence $x \simeq y$, the universally-quantified
variables are abstracted. Our
tatics normalise both sides of the ``potential equivalence'' $x
\simeq y$, and compare the results by syntactic equality. The
difficulty is that the normalisation function needs to consider
variables, constants and operators. For this
reason, we work by reflection. We have one type of reflected
terms for each algebraic structure. The novelty is not the use of reflected
terms, but the use of a type-safe reflection
mechanism where we index the reflected terms by the concrete value that they
represent. Each of these datatype is parametrised over a type \code{c}, which
is the type on which we want to prove equalities (the \emph{carrier}
type). It is also indexed over an implementation of the corresponding interface
for \code{c} (we usually call it \code{p}, because it behaves as a \emph{proof}
telling that the structure \code{c} has the desired properties), indexed over a
context of abstracted variables (a vector $\Gamma$ of \code{n} elements of type
\code{c}). Most importantly, it is indexed over a value of type \code{c},
which is the concrete value being encoded.

@@

\begin{lstlisting}
data ExprG :  Group c $\rightarrow$ (Vect n c) $\rightarrow$ c $\rightarrow$ Type where
    ConstG : (p : Group c) $\rightarrow$ ($\Gamma$:Vect n c) $\rightarrow$ (c1:c) $\rightarrow$ ExprG p $\Gamma$ c1
    PlusG : {p : Group c} $\rightarrow$ {$\Gamma$:Vect n c} $\rightarrow$ {c1:c} $\rightarrow$ {c2:c} 
         $\rightarrow$ ExprG p $\Gamma$ c1 $\rightarrow$ ExprG p $\Gamma$ c2 $\rightarrow$ ExprG p $\Gamma$ (c1+c2)
    MinusG : {p : Group c} $\rightarrow$ {$\Gamma$:Vect n c} $\rightarrow$ {c1:c} $\rightarrow$ {c2:c} 
         $\rightarrow$ ExprG p $\Gamma$ c1 $\rightarrow$ ExprG p $\Gamma$ c2 $\rightarrow$ ExprG p $\Gamma$ (Minus c1 c2)
    NegG : {p : Group c} $\rightarrow$ {$\Gamma$:Vect n c} $\rightarrow$ {c1:c} 
         $\rightarrow$ ExprG p $\Gamma$ c1 $\rightarrow$ ExprG p $\Gamma$ (Neg c1)
    VarG : (p : Group c) $\rightarrow$ {$\Gamma$:Vect n c} 
         $\rightarrow$ (i:Fin n) $\rightarrow$ ExprG p $\Gamma$ (index i $\Gamma$)
\end{lstlisting}

@@

We will explain how to implement the tactic for \code{Group} specifically,
using a normalisation function \code{groupNormalise},
but the other structures are implemented similarly. 
The equivalence we are trying to prove is $x \simeq y$, where $x$ and $y$ are
elements of the type $c$, which implements a group. The reflected term for $x$
is denoted $e_x$, and has type $\code{ExprG}\ p\
\Gamma\ x$, which means that $e_x$ is guaranteed to be the encoding of $x$.  
Similarly,
$y$, is encoded by $e_y$, and its type is indexed by the
value $y$.  Evaluating \code{norm} on $e_x$ produces the
normal form $e_{x'}$ of type $\code{ExprG}\ p\ \Gamma\ x'$ and a proof $p_x$ of
$x \simeq x'$.
Similarly, for $e_y$, it produces
the normal form $e_{y'}$ of type $\code{ExprG}\ p\ \Gamma\ y'$ and a proof
$p_y$ of $y \simeq y'$.  It now suffices to compare $e_{x'}$ and $e_{y'}$
using a syntactic equality test \code{exprG\_eq} on the normal forms.

If the two normalised expressions $e_{x'}$ and $e_{y'}$ are equal, then
they necessarily have the same type\footnote{We work with the
heterogeneous equality by default in Idris, but as always, the only way to
have a proof of \code{a:A = b:B} is when \code{A = B}}, and therefore $x'=y'$, which implies
$x' \simeq y'$ because the propositional equality is more restrictive than an
equivalence relation.  By using the two equivalences $x \simeq x'$ and $y
\simeq y'$, and the new equivalence
$x' \simeq y'$, we can use \code{eq\_preserves\_eq}, to get the desired proof of
$x \simeq y$. We put all of this together in a function
\code{buildProofGroup}:

\begin{lstlisting}
buildProofGroup : (p:Group c) $\rightarrow$ {$\Gamma$:Vect n c} $\rightarrow$ {x : c} $\rightarrow$ {y : c} 
  $\rightarrow$ {x':c} $\rightarrow$ {y':c} $\rightarrow$ (ExprG p $\Gamma$ x') $\rightarrow$ (ExprG p $\Gamma$ y') 
  $\rightarrow$ (x $\simeq$ x') $\rightarrow$ (y $\simeq$ y') $\rightarrow$ (Maybe (x $\simeq$ y))
buildProofGroup p ex' ey' px py with (exprG_eq p ex' ey')
    buildProofGroup p ex' ex' px py | Just refl = 
                Just(eq_preserves_eq x y x' y' lp rp e1_equiv_e2)
    buildProofGroup p ex' ey' px py | Nothing = Nothing
\end{lstlisting}

The argument of type $\code{ExprG}\ p\ \Gamma\ x'$ is the normalised reflected
left hand size of the equivalence, which represents the value $x'$. 
The function also expects proofs of $x \simeq x'$ and
of $y \simeq y'$, which are built by the normalisation process.

@@

Each algebraic structure has a function for reducing reflected terms to normal.
The algebraic structures which concern us admit a canonical representative for
any element, a property which we use to decide equalities. Without this
property, it would be more complicated to decide equivalence without
brute-forcing a series of rewritings, that would have no termination guarantee.

@@

We describe the normal form for rings, which is our most sophisticated
structure. The input to normalisation is an expression 
with sums, products, constants and variables belonging to an
ordered set $\mathcal{V}$ of variables, which is
a polynomial of multiple variables. As output, it produces a
normal form representing the same polynomial. Therefore, we need a
canonical representation of polynomials. 
There are several possibilities, but we choose
classical mathematical conventions: 
the polynomial will be completely developed, i.e., the
distributivity of $*$ over $+$ will be applied until it cannot be applied
further.
This is a simple but effective approach: the benefit of simplicity is that we
can directly produce a proof of equivalence between the new and old indices
during normalisation.
Because the polynomial is completely developed, at the
toplevel, it is a sum:

@@

It may be surprising that the normal form is a sum of product of
monomials, and not directly a sum of monomials. This is because a
monomial is a product of a constant $C_{i}^j$ (e.g. $4$) and of a product of
variables (e.g. $x*y*z$). For example, $5*(x*(y*z))$ is a monomial. Now let's
consider the term $(5*(x*(y*z))) * (4*(z*z))$. This term is not a monomial, but
we could be tempted to simplify it into the monomial $20*(x*(y*(z*(z*z))))$.
However, that would assume that the product is always commutative, which
is not the case for \emph{every} ring.
Therefore, after 
development, the polynomial is a sum of \emph{product of monomials}, and
not directly a sum of monomials. The only rearrangement that can be done
towards the multiplication is to check if two constants are consecutive in a
product, and if so, to replace them by the constant that represents 
their product.

However, because $+$ is always commutative in a ring, the different
products of monomials themselves can be rearranged in different ways in this
sum. This will be done at the lower level for commutative groups if we can
provide an ordering on products of monomials. This ordering will be defined 
with an ordering on monomials, called \code{isBefore\_mon}, which
looks at the order of variables for comparing two monomials
$Monomial_{i}^{j}$ and $Monomial_{i'}^{j'}$.
\[
Monomial_{i}^{j} = C_{i}^{j} * (Var_{i,1}^{j} * \prod_{k=2}^{c} Var_{i,k}^{j})
\text{ and }
Monomial_{i'}^{j'} = C_{i'}^{j'} * (Var_{i',1}^{j'} * \prod_{k=2}^{c'} Var_{i',k}^{j'})
\]

@@

A novelty of our work is that not only have we built a prover for a
specific algebraic structure, but we have also built a hierarchy of provers
for many algebraic structures. Moreover, each prover reuses components
of the others so that
the various simplifications are not duplicated at the different
levels: the normalisation function of each structure uses as much as possible
the normalisation function of the structure from which it inherits. For
example, the normalisation on monoids reuses the normalisation on semi-groups
in such a way that it does not have to deal with associativity. In this
case, the datatype reflecting terms in semi-groups has
the same expressive power as that for monoids. However, 
there is a difficulty with
groups and monoids: normalisation for groups
needs to reuse the normalisation for monoids, meaning that
we will have to encode 
negations somehow, which can't be directly expressed in a monoid. For this
reason, we develop some specific encodings.

@@

We only need to encode negations of variables, as we can simplify 
the negation of a constant into a constant.
Also, there cannot be a negation of something
different non-atomic (i.e. a variable or a constant), because normalisation 
of
groups has systematically propagated \code{Neg} inside the parentheses,
following simplification\footnote{Note that we have to be careful and not
simplify it to $(-a) + (-b)$ as it would assume that $+$ is commutative}
$-(a+b) = -b + -a$.

>>>>

With some additional effort, we can produce a collection of tactics for proving
\emph{equivalences}, rather than only equalities. 
The machinery is very similar and
we gain another
degree of genericity, with the freedom of choosing the equivalence relation
(which can be the usual equality). 
The user can define their own
notion of equivalence, as long as they provide the proofs of the
properties of the relevant algebraic structure. Let's call \code{c} the
\emph{carrier} type, i.e., the type on which we want to prove equivalences. 
The
equivalence relation on \code{c} has the following profile $(\simeq) : \code{c}
\rightarrow \code{c} \rightarrow \code{Type}$\footnote{This \code{Type} would
be a \code{Prop} in systems, like Coq, that make a distinction between the
world of computations and the world of logical statements}, and must be reflexive, symmetric and transitive.

@@

\begin{lstlisting}
interface Set c where
    ($\simeq$) : c $\rightarrow$ c $\rightarrow$ Type
    refl : (x : c) $\rightarrow$ x $\simeq$ x
    sym : {x, y : c} $\rightarrow$ (x $\simeq$ y) $\rightarrow$ (y $\simeq$ x)
    trans : {x, y, z : c} $\rightarrow$ (x $\simeq$ y) $\rightarrow$ (y $\simeq$ z) $\rightarrow$ (x $\simeq$ z)    
    set_eq : (x : c) $\rightarrow$ (y : c) $\rightarrow$ Maybe (x $\simeq$ y)
\end{lstlisting}

@@

Working with equivalences instead of equalities brings one complication : the proofs
of correctness that we produce by hand cannot use Idris' ``rewrite'' mechanism,
which enables rewriting of a subterm by another one, provided that the two
subterms are propositionally equal. 
%
% EB - I think we can skip this detail
%If we need to prove $P\ x$,
%and have a proof of equivalence $pr:x \simeq x'$, we can't use ``rewrite pr" in
%order to transform the goal into $P\ x'$, as we would normally do if the
%available proof would be an equality $x=x'$. 
%
This is a classical problem of
working within a setoid, which can be mitigated by programming
language support for rewriting terms in setoids. However, Idris
is not equipped with any such support. For this reason, we define the 
following lemma, using the methods of the \code{Set} interface:

@@

We have this requirement because we support
any equivalence relation. The user is free to define the equivalence relation of their choice, but it should be compatible with the operations that he is using.
As with \code{Set}, there is no
tactic for \code{Magma}, because there is no property; all
equivalences are again syntactic equivalences, and can thus be proven by
\code{refl}.

@@

\noindent
Examples of magma are \code{Nat} equipped with addition, and \code{List} with
concatenation.  Next, a monoid is a semi-group with the property of neutral element for a distinguished element called \code{Zero}.

@@

When proving an equivalence $x \simeq y$, the universally-quantified
variables are abstracted and they become part of the context. Our
tatics normalise both sides of the ``potential equivalence'' $x
\simeq y$, and compare the results by syntactic equality. The
difficulty is that the normalisation function needs to consider
variables, constants and operators. For this
reason, we work by reflection, which allows us to work on syntax instead of values. We define one type of reflected
terms for each algebraic structure. The novelty is not the use of reflected
terms, but the use of a type-safe reflection
mechanism where we index the reflected terms by the concrete value that they
represent. Each of these datatype is parametrised over a type \code{c}, which
is the type on which we want to prove equalities (the \emph{carrier}
type). It is also indexed over an implementation of the corresponding interface
for \code{c} (we usually call it \code{p}, because it behaves as a \emph{proof}
telling that the structure \code{c} has the desired properties), indexed over a
context of abstracted variables (a vector $\Gamma$ of \code{n} elements of type
\code{c}). Most importantly, it is indexed over a value of type \code{c},
which is the concrete value being encoded.

@@

\begin{lstlisting}
data ExprG :  Group c $\rightarrow$ (Vect n c) $\rightarrow$ c $\rightarrow$ Type where
    ConstG : (p : Group c) $\rightarrow$ ($\Gamma$:Vect n c) $\rightarrow$ (c1:c) $\rightarrow$ ExprG p $\Gamma$ c1
    PlusG : {p : Group c} $\rightarrow$ {$\Gamma$:Vect n c} $\rightarrow$ {c1:c} $\rightarrow$ {c2:c} 
         $\rightarrow$ ExprG p $\Gamma$ c1 $\rightarrow$ ExprG p $\Gamma$ c2 $\rightarrow$ ExprG p $\Gamma$ (c1+c2)
    MinusG : {p : Group c} $\rightarrow$ {$\Gamma$:Vect n c} $\rightarrow$ {c1:c} $\rightarrow$ {c2:c} 
         $\rightarrow$ ExprG p $\Gamma$ c1 $\rightarrow$ ExprG p $\Gamma$ c2 $\rightarrow$ ExprG p $\Gamma$ (Minus c1 c2)
    NegG : {p : Group c} $\rightarrow$ {$\Gamma$:Vect n c} $\rightarrow$ {c1:c} 
         $\rightarrow$ ExprG p $\Gamma$ c1 $\rightarrow$ ExprG p $\Gamma$ (Neg c1)
    VarG : (p : Group c) $\rightarrow$ ($\Gamma$:Vect n c) 
         $\rightarrow$ (i:Fin n) $\rightarrow$ ExprG p $\Gamma$ (index i $\Gamma$)
\end{lstlisting}

@@

We will explain how to implement the tactic for \code{Group} specifically,
but the other structures are implemented similarly.
The equivalence we are trying to prove is $x \simeq y$, where $x$ and $y$ are
elements of the type $c$, which implements a group. The reflected term for $x$
is denoted $e_x$, and has type $\code{ExprG}\ p\
\Gamma\ x$, which means that $e_x$ is guaranteed to be the encoding of $x$.  
Similarly,
$y$, is encoded by $e_y$, and its type is indexed by the
value $y$.  Evaluating \code{norm} on $e_x$ produces the
normal form $e_{x'}$ of type $\code{ExprG}\ p\ \Gamma\ x'$ and a proof $p_x$ of
$x \simeq x'$.
Similarly, for $e_y$, it produces
the normal form $e_{y'}$ of type $\code{ExprG}\ p\ \Gamma\ y'$ and a proof
$p_y$ of $y \simeq y'$.  It now suffices to compare the normal forms $e_{x'}$ and $e_{y'}$
using a syntactic equality test \code{exprG\_eq}. If the two normalised expressions $e_{x'}$ and $e_{y'}$ are equal, then
they necessarily have the same type\footnote{We work with the
heterogeneous equality by default in Idris, but as always, the only way to
have a proof of \code{a:A = b:B} is when \code{A = B}}, and therefore $x'=y'$, which implies
$x' \simeq y'$ because the propositional equality is more restrictive than an
equivalence relation.  By using the two equivalences $x \simeq x'$ and $y
\simeq y'$, and the new equivalence
$x' \simeq y'$, we can use \code{eq\_preserves\_eq}, to get the desired proof of
$x \simeq y$. We put all of this together in a function
\code{buildProofGroup}:

\begin{lstlisting}
buildProofGroup : (p:Group c) $\rightarrow$ {$\Gamma$:Vect n c} $\rightarrow$ {x : c} $\rightarrow$ {y : c} 
  $\rightarrow$ {x':c} $\rightarrow$ {y':c} $\rightarrow$ (ExprG p $\Gamma$ x') $\rightarrow$ (ExprG p $\Gamma$ y') 
  $\rightarrow$ (x $\simeq$ x') $\rightarrow$ (y $\simeq$ y') $\rightarrow$ (Maybe (x $\simeq$ y))
buildProofGroup p ex' ey' px py with (exprG_eq p ex' ey')
    buildProofGroup p ex' ey' px py | Just ex'_equiv_ey' = 
                Just(eq_preserves_eq x y x' y' px py ex'_equiv_ey')
    buildProofGroup p ex' ey' px py | Nothing = Nothing
\end{lstlisting}

The arguments of type $\code{ExprG}\ p\ \Gamma\ x'$ and $\code{ExprG}\ p\ \Gamma\ y'$ are the normalised reflected
left and right hand sides of the equivalence, which respectively represent the value $x'$ and $y'$. 
This function also expects proofs of $x \simeq x'$ and
of $y \simeq y'$, which are built by the normalisation process.

@@

Each algebraic structure is equipped with a function for reducing reflected terms to their normal form.
The algebraic theories which concern us admit a canonical representative\footnote{This only holds in the absence of additional axioms}  for
any element, a property which we use to decide equalities. Without this
property, it would be more complicated to decide equivalence without
brute-forcing a series of rewritings, that would have no termination guarantee.

@@

We describe the normal form for rings, which is our most sophisticated
structure. The input to the normalisation function is an expression 
with sums, products, constants and variables belonging to an
ordered set $\mathcal{V}$ of variables. In short, the normalisation function takes in input a polynomial of multiple variables. As output, it produces a
normal form representing the same polynomial. Therefore, we need a
canonical representation of polynomials. 
There are several possibilities, but we choose
classical mathematical conventions: 
the polynomial will be completely developed, i.e., the
distributivity of $*$ over $+$ will be applied until it cannot be applied
further.
This is a simple but effective approach: the benefit of simplicity is that we
can directly produce a proof of equivalence between the new and old concrete values
during normalisation.
Because the polynomial is completely developed, at the
toplevel, it is a sum:

@@

It may be surprising that the normal form is a sum of product of
monomials, and not directly a sum of monomials. This is because a
monomial is a product of a constant $C_{i}^j$ (e.g. $5$) and of a product of
variables (e.g. $x*y*z$). For example, $5*(x*(y*z))$ is a monomial. Now let's
consider the term $(5*(x*(y*z))) * (4*(z*z))$. This term is not a monomial, but
we could be tempted to simplify it into the monomial $20*(x*(y*(z*(z*z))))$.
However, that would assume that the product is always commutative, which
is not the case for \emph{every} ring.
Therefore, after 
development, the polynomial is a sum of \emph{product of monomials}, and
not directly a sum of monomials. The only rearrangement that can and needs to be done
towards the multiplication is to check if two constants are consecutive in a
product, and if so, to replace them by the constant that represents 
their product.

However, because $+$ is always commutative in a ring, the different
products of monomials themselves can be rearranged in different ways in this
sum. This will be done at the lower level for commutative groups if we can
provide an ordering on products of monomials. This ordering will be defined by using an ordering on monomials, called \code{isBefore\_mon}, which
looks at the order of variables for comparing two monomials
$Monomial_{i}^{j}$ and $Monomial_{i'}^{j'}$.
\[
Monomial_{i}^{j} = C_{i}^{j} * (Var_{i,1}^{j} * \prod_{k=2}^{c} Var_{i,k}^{j})
\text{ and }
Monomial_{i'}^{j'} = C_{i'}^{j'} * (Var_{i',1}^{j'} * \prod_{k=2}^{c'} Var_{i',k}^{j'})
\]

@@

A novelty of our work is that instead of building a prover for a
specific algebraic structure (like ring), we have built a hierarchy of provers
for many algebraic structures. Moreover, each prover reuses components
of the others so that
the various simplifications are not duplicated at the different
levels: the normalisation function of each structure uses as much as possible
the normalisation function of the structure from which it inherits. For
example, the normalisation on monoids reuses the normalisation on semi-groups
in such a way that it does not have to deal with associativity. In this
case, the datatype reflecting terms in semi-groups has
the same expressive power as that for monoids, so a term of a monoid can easily be transformed into a term of a semi-group. However, 
there is a difficulty with
groups and monoids: normalisation for groups
needs to reuse the normalisation for monoids, meaning that
we will have to encode 
negations somehow, which can't be directly expressed in a monoid. For this
reason, we develop some specific encodings.

@@

We only need to encode negations of variables, as we can simplify 
the negation of a constant into a constant.
Also, there cannot be a negation of something
different non-atomic (i.e. a term that is not a variable or a constant), because normalisation 
of
groups has systematically propagated \code{Neg} inside the parentheses,
following simplification\footnote{Note that we have to be careful and not
simplify it to $(-a) + (-b)$ as it would assume that $+$ is commutative}
$-(a+b) = -b + -a$.

### 2017-05-19 paper-CICM-17/paper.tex
Pruning up to 2.3; add acknowledgements
Some additional changes

\begin{abstract}
One difficulty with programming with dependent types is that proof obligations
arise naturally once programs become even moderately sized. For example,
implementing an adder for binary numbers indexed over their natural number
equivalents naturally leads to proof obligations for equalities of expressions
over natural numbers. The need for these equality proofs comes, in intensional
type theories, from the fact that the propositional equality enables us to
prove as equal terms that are not judgementally equal, which means that the
typechecker can't always obtain equalities by reduction.  As far as possible,
we would like to solve such proof obligations automatically. In this paper, we
show one way to automate these proofs by reflection in the dependently typed
programming language Idris.  We show how defining reflected terms indexed by
the original Idris expression allows us to construct and manipulate proofs. We
build a hierarchy of tactics for proving equivalences in semi-groups, monoids,
commutative monoids, groups, commutative groups, semi-rings and rings. We also
show how each tactic reuses those from simpler structures, thus avoiding
duplication of code and proofs.

@@

>>>>

\begin{abstract}
One difficulty with reasoning and programming with dependent types is that proof obligations arise naturally once programs become even moderately sized. For example,
implementing an adder for binary numbers indexed over their natural number
equivalents naturally leads to proof obligations for equalities of expressions
over natural numbers. The need for these equality proofs comes, in intensional
type theories, from the fact that the propositional equality enables us to
prove as equal terms that are not judgementally equal, which means that the
typechecker can't always obtain equalities by reduction.  As far as possible,
we would like to solve such proof obligations automatically. In this paper, we
show one way to automate these proofs by reflection in the dependently typed
programming language Idris.  We show how defining reflected terms indexed by
the original Idris expression allows us to construct and manipulate proofs. We
build a hierarchy of tactics for proving equivalences in semi-groups, monoids,
commutative monoids, groups, commutative groups, semi-rings and rings. We also
show how each tactic reuses those from simpler structures, thus avoiding
duplication of code and proofs.

@@

\subsubsection*{Acknowledgements}
We thank the anonymous reviewers and Jacques Carette for their insightful
comments on an earlier draft. We are also grateful for the support of
the Scottish Informatics and Computer Science Alliance (SICSA) and
EPSRC grant EP/N024222/1.
